---
title: "Snapshot of Tropical Storms"
author: "Leap Kasten & Isabel Zamudio-Moore"
date: "Last updated on `r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 2
    toc_float: true
    df_print: kable
---

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Set R Markdown chunk defaults:
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE, 
  fig.width = 16/2, fig.height = 9/2
)
```

<b> INTRODUCTION: </b>

There are many kinds of storms that draw their power from the wind. Tropical storms are among them. These storms, including hurricanes, are often associated with the damage caused by the waves they create. Hurricanes are wind-based funnel storms that form with low pressure and create storm clouds by picking up moisture from the ocean^[Dunbar, Brian. “What Are Hurricanes?” NASA, NASA, 13 May 2015, http://www.nasa.gov/audience/forstudents/k-4/stories/nasa-knows/what-are-hurricanes-k4.html.]. We decided to investigate the extreme winds of tropical storms from 2000-2015 using NOAA hurricane data^[NOAA. “Hurricanes and Typhoons, 1851-2014.” RSNA Pneumonia Detection Challenge | Kaggle, 20 Jan. 2017, http://www.kaggle.com/noaa/hurricane-database .]we found on Kaggle. 

<!--
Loading Packages & Data
-->
```{r}
# Load Packages
library(tidyverse)
library(lubridate)
library(stringr)
library(tidyselect)
library(leaflet)

# Load Data
pacific <- read.csv("pacific.csv")
atlantic <- read.csv("atlantic.csv")
```


<!--
Hurricane Data
-->
```{R}
# Clean Variables of Interest
# Change Name to a character and Max. Wind to a number
atlantic <- atlantic %>%
  mutate(Name=as.character(Name))

# Change Name to a character and Max. Wind to a number
pacific<- pacific %>%
  mutate(Name=as.character(Name))

# Join the datasets for Hurricane
hurricane <- pacific %>%
  full_join(atlantic, by= c("Name", "Date", "Maximum.Wind")) %>%
  select(Name, Date, Maximum.Wind)

# Filter to include dates between 2000-2015 
# Formate Date & Create Event
hurricane2<- hurricane %>%
  filter( 20160101 >= Date & Date> 20000101)%>%
  mutate(Date=ymd(Date), Event= "Tropical Cyclone")

# Group Tropical Storms by name, and find average maximum wind speed in descending order
hurricane3 <- hurricane2 %>%
  group_by(Name) %>%
  summarize(MW = mean(Maximum.Wind))%>%
  arrange(desc(MW))%>%
  slice(1:15)
```


<b> Wind Speeds Over Time: </b>

```{r}
# Create Scatter Plot
ggplot(data= hurricane2, mapping= aes(x=Date, y= Maximum.Wind, color=Maximum.Wind))+
  geom_point()+
  labs(title= "Tropical Storm's Average Maximum Wind Speeds in 15 Years", x= "Date", y= "Average Maximum Wind Speed")
```

Are There Trends in Tropical Storms' Average Maximum Wind Speeds?

  The chart above, shows the tropical storms’ averaged maximum wind speed using the atlantic and pacific datasets^[NOAA. “Hurricanes and Typhoons, 1851-2014.” RSNA Pneumonia Detection Challenge | Kaggle, National Hurricane Center, 20 Jan. 2017, http://www.kaggle.com/noaa/hurricane-database.] from NOAA. Looking at the tropical storm data globally over the 15-year period we made some observations which can be explained by previously known facts about tropical storms. For instance, the chart made with the datasets ^[NOAA. “Hurricanes and Typhoons, 1851-2014.” RSNA Pneumonia Detection Challenge | Kaggle, National Hurricane Center, 20 Jan. 2017, http://www.kaggle.com/noaa/hurricane-database.] from NOAA, shows a fairly consistent pattern of plumes separated by gaps of time between observations. These gaps can be explained by the fact that tropical cyclones only occur from June to November^[NOAA. “Tropical Cyclone Climatology.” Hurricane Preparedness - Hazards, NOAA's National Weather Service, http://www.nhc.noaa.gov/climo/ . ].
    One may note that in 2013 the range of average maximum wind speed was much smaller than the surrounding years; this could be attributed to the fact the Pacific and Atlantic hurricane seasons were both shorter than average in 2013^[NOAA National Centers for Environmental Information, State of the Climate: Hurricanes and Tropical Storms for Annual 2013, published online January 2014, retrieved on December 19, 2018 from https://www.ncdc.noaa.gov/sotc/tropical-cyclones/201313.].  Another year of interest is 2015 because of its two storms with ridiculously high wind speeds, which can be noted on the chart made using NOAA’s dataset^[NOAA. “Hurricanes and Typhoons, 1851-2014.” RSNA Pneumonia Detection Challenge | Kaggle, National Hurricane Center, 20 Jan. 2017, http://www.kaggle.com/noaa/hurricane-database.]. Hurricanes Danny and Joaquin were both huge category four hurricanes in the Atlantic that year; Danny is a notable storm because it only lasted for a few days before dissipating whereas Hurricane Joaquin spent nearly a month sweeping up the gulf coast^[Dolace, Chris, and Quincy Vagell. “2015 Hurricane Season in Review: 11 Things We Will Remember.” The Weather Channel, The Weather Channel, 2015, https://weather.com/storms/hurricane/news/hurricane-season-2015-recap/.].
